Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2016-08-30 LaTeX/Turnover.tex
Intro improved, not yet finished

Animals and angiosperms exhibit extremely diverse sex determination systems, as reviewed by \citet{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric/dioecious clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex determination systems can be found the same species, including cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.

\noindent
Depending on the prominence of transitions to ESD in the manuscript include something like (currently quoted):

``Transitions have repeatedly occurred between environmental sex determination and genotypic sex determination, as exemplified by the distribution of temperature sex determination among reptiles: either temperature or genes provide the initial trigger in closely related species (Ewert and Nelson 1991; Pokorna and Kratochvil 2009; Ezaz et al. 2009) or even conspecific populations (Pen et al. 2010). Similar situations are found in fishes (e.g., Conover and Heins 1978a).''

@@

\noindent
\textcolor{red}{Brief description of sex ratio adjustment and sexual antagonism theories:}

@@

Sex ratio selection might be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
For example, feminizing mutations may invade when female biased sex ratios are favoured due to interdemic selection \citep{Wilson:1981vm,Vuillleumier:2007bh}.
In other cases, flexible sex determination systems may be favoured in order to exploit environmental conditions that are optimal for males or females, creating locally biased sex ratios \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In other situations, sex ratio selection may favour transitions in order to restore equal sex ratios. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., driving X or Y chromosomes are inherited disproportionately often), which leads to biased sex ratios. They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations) can then spread, restoring an even sex ratio.

\noindent
\textcolor{red}{We add haploid selection:}

Here, we use mathematical models to find the conditions under which new sex determination systems are favoured by selection where we include a period of selection among haploid gametes/gametophytes. 
{\color{blue}
FROM PREVIOUS PAPER:
In plants, selection among haploid male gametophytes is thought to be pervasive \cite{SKOGSMYR:2002ce,Moore:2011jt,Marshall:2016fe}; in \textit{Arabidopsis}, 60-70\% of all genes are expressed during the haploid phase \cite{Borg:2009jpa}, and pollen expressed genes exhibit stronger signatures of purifying selection and positive selection \cite{Arunkumar:2013iq,Gossmann:2014dua}.
For agricultural breeding, pollen has been exposed to a variety of selection pressures \textit{in vivo} and \textit{in vitro}, including temperature \cite{Hedhly:2004iv,Clarke:2004ir}, herbicides \cite{Frascaroli:2001ee}, metals \cite{Searcy:1985vt}, water stress \cite{Ravikumar:2003uo}, and pathogens \cite{Ravikumar:2012ej}, resulting in an increased frequency of resistant genotypes among the diploid sporophytic offspring. 
In animals, expression during the haploid sperm stage is traditionally thought to be suppressed \cite{Hecht:1998hz}, although recent evidence suggests that the extent and selective importance of postmeiotic gene expression may be underestimated \cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et,Immler:2014im}. 
}

Here, suggest that the canonical view (no haploid expression in animals, genome highly chromatinized in sperm and not expressed) might be based on model organisms, such as mice, where sperm is sufficiently short-lived that transcripts provisioned during spermatogenesis may be sufficient without further haploid transcription (although note that the Vibranovski lab results are in mice showing some transcription does occur). 
In broadcast spawning animal species (e.g., corals, many fish) and species where sperm typically requires greater longevity, expression of the haploid genotype may be required (Immler paper indicates this, but not that strongly - as I remember). 
We can use this suggestion in discussion to speculate in what species the processes we study might be looked for (i.e.., animals with multiple matings, broadcast spawning and/or long-lived sperm and outcrossing/non-pollen-limited plants).

{\color{blue} 
\noindent FROM PREVIOUS PAPER:

The maintenance of polymorphism at loci that experience sex specific selection in both haploid and diploid phases was considered by Immler et al. \cite{Immler:2012tl}, demonstrating that polymorphisms can be maintained by sexually antagonistic selection or overdominance as well as by conflicting selection pressures in haploids and diploids (haploid-diploid conflict or ploidally antagonistic selection) or a combination of these selective regimes.

We add haploid selection (and justification, see below)
\\
Also discuss the fact that, in terms of recombination suppression, haploid selection among male gametes generates selection pressure similar to that of male specific selection. 
\\
What will be the result: where there is sex biases and sex-specific selection

@@

@@

DRAFT (improve): In \citet{Ubeda:2015fx}, the new sex determining locus spreads because it arises in linkage with a locus that experiences drive. They assume that drive occurs predominantly in one sex, e.g., during spermatogenesis or a 'killer' sperm. A driving allele is maintained at an intermediate frequency by selection, e.g., because it causes male sterility when homozygous (because all male sperm are killed). Y chromosomes that arise in linkage with the driving allele spread because they allow drive to occur more often, thus genetic sex determination with a sex ratio bias evolves. 
Thus \citet{Ubeda:2015fx} also find that genetic sex determiners can invade, despite causing sex ratios to become biased. 
Finally, they show that autosomal 'restorers' that negate the effects of meiotic drive can invade and restore an equal sex ratio. 
When sex ratio bias occurs due to haploid selection, a natural class of sex ratio `restorers' exist because haploid selection often occurs in a context that is determined by the diploid parents. 
For example, the intensity of pollen competition can be manipulated by altering style length \citep{Travers:2001,Lankinen:2001gc,Ruane:2009vt}, delaying stigma receptivity \citep{Galen:1986wq,Lankinen:2011if} and/or delaying pollen tube growth in the pistil \citep{Herrero:2003jf}. 
Where the X and Y have fitness differences, \citet{Hough:2013uo} and \citet{Otto:2015va} demonstrated that mothers should generally evolve to balance sex ratios by reducing the intensity of haploid competition.

@@

{\color{blue}
FROM THESIS:
Generally, any sex-linked gene that harbours genetic variation in haploid fitness should cause sex ratios to become biased. 
Sex ratio bias caused by pollen competition has previously been discussed in the context of Y-linked deleterious mutations, which are thought to build up after recombination suppression evolves \citep{Lloyd:1974tz,Stehlik:2005ul}. 
Sex ratios can also become biased due to meiotic drive;
in a classic paper, \citet{Hamilton:1967ts} showed that X- or Y-linked alleles that experience meiotic drive will bias sex ratios. 
He assumed that driving alleles are under directional selection and spread to fixation but such alleles can also be maintained at intermediate frequencies by selection \citep{Feldman:1989gm,Holman:2015en}. 
When sex ratios are biased, other loci are expected to evolve to restore equal sex ratios. 
Indeed, alleles that negate the effect of sex-linked meiotic drivers and restore equal sex ratios have been identified \citep{Stalker:1961th,Smith:1975ft}. 
A similar process occurs with cytoplasmic male sterility alleles (that cause biased sex ratios) and nuclear `restorer' genotypes \citep{Frank:1989vl}.

Several aspects of the relationship between haploid selection (e.g., pollen or sperm competition) and sex ratios remain to be explored. 
For example, new sex-determining systems (particularly transitions between male and female heterogamety) can be favoured in order to restore equal sex ratios in populations that have a sex ratio bias \citep{Bull:1983vi,Kozielska:2010vm,Ubeda:2015fx}.
Based on the results of Chapter \ref{ch:sexchromosomes}, we would expect that sex ratio biases would occur via associations between sex-determining loci and loci that experience haploid selection. 
However, these associations should also select against transitions between sex-determining systems, as has been found with sexually antagonistic selection \citep{vanDoorn:2007eu,vanDoorn:2010hu}. 
It is not clear how the spread of new sex determination systems would be influenced by the combination of sex ratio biases and favourable associations between haploid selected loci and sex-determining regions. 
Finally, \citet{Hamilton:1967ts} pointed out that biased sex ratios can affect population size because the number of offspring in each generation is typically determined by the number of females. 
Population density can, in turn, affect the intensity of pollen/sperm competition in future generations because fewer males are available to donate pollen/sperm in a particular area. 
Thus, a feedback could occur between population densities and haploid selection, which has not yet been investigated. 
}

>>>>

Animals and angiosperms exhibit extremely diverse sex determination systems, as reviewed by \citet{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric/dioecious clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex determination systems can be found the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.
In addition, multiple transitions have occurred between genetic and environmental sex determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

%\textcolor{red}{Brief description of sex ratio adjustment and sexual antagonism theories:}

@@

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
For example, flexible sex determination systems may be favoured in order to exploit local environmental conditions that are optimal for males or females, which creates locally biased sex ratios \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In addition, feminizing mutations may invade when female biased sex ratios are favoured due to selection among demes \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
In other situations, sex ratio selection may favour transitions in order to restore equal sex ratios. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, restoring an even sex ratio.

%\textcolor{red}{We add haploid selection:}

Here, we use mathematical models to find the conditions under which new sex determination systems are favoured by selection where we include a period of selection among haploid gametes/gametophytes. 
Selection among haploid genotypes is thought to occur primarily among pollen/sperm, which can compete whenever there are more pollen/sperm than required for fertilization \citep{Mulcahy:1996ha,JOSEPH:2004haa}. 
Haploid selection may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

%Also discuss the fact that, in terms of recombination suppression, haploid selection among male gametes generates selection pressure similar to that of male specific selection. 
%And sex ratio bias

%What will be the result: where there is sex biases and sex-specific selection

@@

{\color{blue} 
FOR RESULTS?

FROM PREVIOUS PAPER:
The maintenance of polymorphism at loci that experience sex specific selection in both haploid and diploid phases was considered by Immler et al. \cite{Immler:2012tl}, demonstrating that polymorphisms can be maintained by sexually antagonistic selection or overdominance as well as by conflicting selection pressures in haploids and diploids (haploid-diploid conflict or ploidally antagonistic selection) or a combination of these selective regimes.  
}

@@

DRAFT (improve): In \citet{Ubeda:2015fx}, the new sex determining locus spreads because it arises in linkage with a locus that experiences drive. They assume that drive occurs predominantly in one sex, e.g., during spermatogenesis or a 'killer' sperm. A driving allele is maintained at an intermediate frequency by selection, e.g., because it causes male sterility when homozygous (because all male sperm are killed). Y chromosomes that arise in linkage with the driving allele spread because they allow drive to occur more often, thus genetic sex determination with a sex ratio bias evolves. 
Thus \citet{Ubeda:2015fx} also find that genetic sex determiners can invade, despite causing sex ratios to become biased. 
Finally, they show that autosomal 'restorers' that negate the effects of meiotic drive can invade and restore an equal sex ratio. 
OTHER RESTORERS.
\textcolor{blue}{Indeed, alleles that negate the effect of sex-linked meiotic drivers and restore equal sex ratios have been identified \citep{Stalker:1961th,Smith:1975ft}. 
A similar process occurs with cytoplasmic male sterility alleles (that cause biased sex ratios) and nuclear `restorer' genotypes \citep{Frank:1989vl}. 
}
When sex ratio bias occurs due to haploid selection, a natural class of sex ratio `restorers' exist because haploid selection often occurs in a context that is determined by the diploid parents. 
For example, the intensity of pollen competition can be manipulated by altering style length \citep{Travers:2001,Lankinen:2001gc,Ruane:2009vt}, delaying stigma receptivity \citep{Galen:1986wq,Lankinen:2011if} and/or delaying pollen tube growth in the pistil \citep{Herrero:2003jf}. 
Where the X and Y have fitness differences, \citet{Hough:2013uo} and \citet{Otto:2015va} demonstrated that mothers should generally evolve to balance sex ratios by reducing the intensity of haploid competition.

@@

\textcolor{red}{\noindent Discuss patterns that might be looked for:}

In broadcast spawning animal species (e.g., corals, many fish) and species where sperm typically requires greater longevity, haploid selection may be stronger because transcripts shared during spermatogenesis may become depleted \citep{Immler:2014im}. 
\textcolor{red}{also, mating systems (e.g., fewer alleles are available during haploid competition in monogamous species), selfing rates, and estimates of pollen limitation could be used as indicators of the intensity of haploid selection}

{\color{blue}
Several aspects of the relationship between haploid selection (e.g., pollen or sperm competition) and sex ratios remain to be explored. 
For example, new sex-determining systems (particularly transitions between male and female heterogamety) can be favoured in order to restore equal sex ratios in populations that have a sex ratio bias \citep{Bull:1983vi,Kozielska:2010vm,Ubeda:2015fx}.
Based on the results of Chapter \ref{ch:sexchromosomes}, we would expect that sex ratio biases would occur via associations between sex-determining loci and loci that experience haploid selection. 
However, these associations should also select against transitions between sex-determining systems, as has been found with sexually antagonistic selection \citep{vanDoorn:2007eu,vanDoorn:2010hu}. 
It is not clear how the spread of new sex determination systems would be influenced by the combination of sex ratio biases and favourable associations between haploid selected loci and sex-determining regions. 
Finally, \citet{Hamilton:1967ts} pointed out that biased sex ratios can affect population size because the number of offspring in each generation is typically determined by the number of females. 
Population density can, in turn, affect the intensity of pollen/sperm competition in future generations because fewer males are available to donate pollen/sperm in a particular area. 
Thus, a feedback could occur between population densities and haploid selection, which has not yet been investigated. 
}
